The Mass Distribution of Stellar-Mass Black Holes

### 2010-11-05 paper.tex
Final changes before ApJ.

\begin{abstract}
  We perform a Bayesian analysis of the mass distribution of
  stellar-mass black holes using the observed masses of 15 low-mass
  X-ray binary systems undergoing Roche lobe overflow and five
  higher-mass, wind-fed X-ray binary systems.  Using Markov Chain
  Monte Carlo calculations, we model the mass distribution both
  parametrically---as a power law, exponential, gaussian, combination
  of two gaussians, or log-normal distribution---and
  non-parametrically---as histograms with varying numbers of bins.  We
  provide confidence bounds on the shape of the mass distribution in
  the context of each model and compare the models with each other by
  calculating their relative Bayesian evidence as supported by the
  measurements, \changed{taking into account the number of degrees of
    freedom of each model.}  The mass distribution of the low-mass
  systems is best fit by a power-law, while the distribution of the
  complete sample is best fit by the exponential model.  \changed{This
    difference clearly indicates that the character of the low mass
    X-ray binary systems differs from the higher-mass systems.}  We
  examine the existence of a ``gap'' between the most massive neutron
  stars and least massive black holes.  \changed{Our analysis
    generates distributions of lower bounds on the black hole mass
    distribution; we use the 1\% quantile of the mass distribution,
    $M_{1\%}$, for a lower bound because some of our distributions do
    not have a sharp lower limit.  For the best model (the power law)
    fitted to the low-mass systems this distribution of lower-bounds
    has $M_{1\%} > 4.3$ $\Msun$ with 90\% confidence, while the best
    model (the exponential) fitted to all 20 systems has $M_{1\%} >
    4.5$ $\Msun$ with 90\% confidence.}  We conclude that our sample
  of black hole masses provides strong evidence of a gap between the
  maximum neutron star mass and \changed{the lower bound on} black
  hole masses.  \changed{Our results on the low-mass sample are in
    qualitative agreement with those of \citet{Ozel2010}, although our
    broad model selection analysis more reliably reveals the best-fit
    quantitative description of the underlying mass distribution.  The
    results on the complete sample of low- and higher-mass systems are
    in qualitative agreement with \citet{Fryer2001} although the
    presence of a mass gap remains theoretically unexplained.}
\end{abstract}

@@

Observations of low-mass X-ray binaries in both the optical and X-ray
can provide a measurement of the mass of the compact object in these
systems.  \changed{The current sample of stellar mass black holes with
  dynamically measured masses includes 15 low-mass, Roche lobe
  overflowing systems and 5 wind-fed higher-mass systems.  Hence,
  sophisticated statistical analyses of the black hole mass
  distribution in these systems are possible.}

The first study of the mass distribution of stellar-mass black holes,
in \citet{Bailyn1998}, examined a sample of seven low-mass X-ray
binaries thought to contain a black hole, concluding in a Bayesian
analysis that the mass function was strongly peaked around seven solar
masses.  \changed{(A similar analysis of the neutron star mass
  distribution can be found in \citet{Finn1994}.)}  \citet{Bailyn1998}
found evidence of a ``gap'' between the least massive black hole and a
``safe'' upper limit for neutron star masses of 3 $\Msun$ (e.g.\
\citet{Kalogera1996}).  Such a gap is puzzling in light of theoretical
studies that predict a continuous distribution of compact object
supernova remnant masses with a smooth transition from neutron stars
to black holes \citep{Fryer2001}.  \changed{(But note that
  \citet{Fryer2001} considered binary evolution effects only
  heuristically and put forward some possible explanations for the gap
  from \citet{Bailyn1998} both in the context of selection effects or
  in connection to the energetics of supernova explosions.)}

\changed{Towards the end of our analysis work, we became aware of a
  more recent study \citep{Ozel2010}, also in a Bayesian framework,
  analyzing the low-mass X-ray binary sample.}  It largely confirmed
these findings, examining 16 low-mass X-ray binary systems containing
black holes and finding a strongly peaked distribution at $7.8 \pm 1.2
\, \Msun$.  \citet{Ozel2010} use two models for the mass function: a
Gaussian and a decaying exponential with a minimum ``turn-on'' mass
(motivated by the analytical model of the black-hole mass function in
\citet{Fryer2001}).  \changed{\citet{Ozel2010} do not provide
  confidence limits for the minimum black hole mass, instead
  discussing only the model parameters at the peak of their posterior
  distributions.  They also do not perform any model selection
  analysis; thus, they give the distribution of parameters within each
  of their models, but cannot say which model is more likely to
  correspond to the true distribution of black hole masses.}
Nevertheless, it appears that their analysis confirms the existence of
a mass gap.  \citet{Ozel2010} discuss possible selection effects that
could lead to the appearance of a mass gap, but conclude these effects
could not produce the observed gap, which they therefore claim is a
real property of the black hole mass distribution.

In this work, we use observations of 20 galactic X-ray binaries that
contain black holes to constrain the mass function for stellar mass
black holes.  We consider separately 15 low-mass X-ray binaries
containing Roche-lobe-filling secondaries with masses smaller than the
black hole primary, and a sample including these systems and five
higher-mass, wind-fed systems where the secondary is heavier than the
black hole primary.

@@

\changed{We define the ``minimum black hole mass'' to be the 1\%
  quantile, $M_{1\%}$, in the black hole mass distribution (see
  Section \ref{sec:minimum-mass}).  In agreement with \citet{Ozel2010}
  and \citet{Bailyn1998}, we find strong evidence for a mass gap among
  the best models for both samples.  Our analysis gives distributions
  for $M_{1\%}$ implied by the data in the context of each of our
  models for the black hole mass distribution.  In the context of the
  best model for the low-mass systems (a power-law), the distribution
  for $M_{1\%}$ has $M_{1\%} > 4.3$ $\Msun$ (at 90\% confidence); in
  the context of the best model for the combined sample of lower- and
  high-mass systems the distribution of $M_{1\%}$ has $M_{1\%} > 4.5$
  $\Msun$ with 90\% confidence.  Further, in the context of models
  with lower evidence, most also have a mass gap, with 90\% confidence
  bounds on $M_{1\%}$ significantly above a ``safe'' maximum neutron
  star mass of 3 $\Msun$ \citep{Kalogera1996}.}

\changed{We find that, for the low-mass X-ray binary sample, the
  theoretical model from \citet{Fryer2001}, a decaying exponential is
  strongly disfavored by our model selection.  We find that the
  low-mass systems are best described by a power law, followed
  closely by a Gaussian (which is the other model from
  \citet{Ozel2010}).  On the other hand, we find that the theoretical
  model from \citet{Fryer2001} is the preferred model for the combined
  sample of low- and high-mass X-ray binaries.  A model with two
  separate Gaussian peaks also has relatively high evidence for the
  combined sample of systems.  The difference in best-fitting model
  clearly indicates that the character of the low mass X-ray binary
  systems differs from the higher-mass systems.}

\changed{The structure of this paper is as follows.  In Section
  \ref{sec:systems} we discuss the 15 systems that comprise the
  low-mass X-ray binary black hole sample and the 5 additional
  higher-mass, wind-fed systems that make up the complete sample.  In
  Section \ref{sec:models} we discuss the Bayesian techniques we use
  to analyze the black hole mass distribution, the techniques we use
  for model selection, and the parametric and non-parametric models we
  will for the black hole mass distribution.  In Section
  \ref{sec:results} we discuss the results of our analysis and model
  selection.  In Section \ref{sec:minimum-mass} we discuss the
  distribution of the minimum black hole mass implied by the analysis
  of Section \ref{sec:results}.  In Section \ref{sec:conclusion} we
  summarize our results and comment on the significance of the
  observed mass gap in the context of theoretical models.  Appendix
  \ref{sec:mcmc} describes MCMC techniques in some detail.  Appendix
  \ref{sec:reversible-jump-mcmc} explains our novel algorithm for
  efficiently performing the reversible jump MCMC computations used in
  the model comparisons of Section \ref{sec:results} (but see also
  \citet{Farr2010}).}

@@

The 20 X-ray binary systems on which this study is based are listed in
Table \ref{tab:sources}.  We separate the systems into 15 low-mass
systems in which the central black hole appears to be fed by
roche-lobe overflow from the secondary, and 5 higher-mass systems in
which the black hole is fed via winds (these systems all have a
secondary that appears to be more massive than the black hole).  The
low- and higher-mass systems undoubtedly have different evolutionary
tracks, and therefore it is reasonable that they would have different
black-hole mass distributions.  We will first analyze the 15 low-mass
systems alone (Section \ref{sec:results-low-mass}), and then the
complete sample of 20 systems (Section \ref{sec:higher-mass}).

@@

In this section we describe the statistical analysis we will apply to
various models for the underlying mass distribution from which the
low-mass sample and the complete sample of X-ray binary systems in
Table \ref{tab:sources} were drawn.  The results of our analysis are
presented in Section \ref{sec:results}.

@@

We consider histograms with up to five bins in this work.  We will see
that the evidence for the histogram models (see Sections
\ref{sec:model-selection}, \ref{sec:low-mass-model-selection}, and
\ref{sec:high-mass-model-selection}) from both the low-mass and
complete datasets is decreasing as the number of bins reaches five,
indicating that increasing the number of bins beyond five would not
sufficiently improve the fit to the mass distribution to compensate
for the extra parameter-space volume implied by the additional
parameters.

@@

In this section we discuss the results of our MCMC analysis of the
posterior distributions of parameters for the models in Sections
\ref{sec:parametric-models} and \ref{sec:non-parametric-models}.  We
also discuss model selection results.  The results in Section
\ref{sec:results-low-mass} apply to the low-mass sample of
systems, while those of Section \ref{sec:higher-mass} apply to the
complete sample of systems.

@@

\begin{figure}
  \begin{center}
    \plotone{plots/dist}
  \end{center}
  \caption{\label{fig:dists} The median (solid line), 10\% (lower
    dashed line) and 90\% (upper dashed line) values of the black hole
    mass distribution, $p(M|\theta)$, at various masses implied by the
    posterior $p(\theta|d)$ for the models discussed in Sections
    \ref{sec:parametric-models} and \ref{sec:non-parametric-models}.
    These distributions use only the 15 low-mass observations in Table
    \ref{tab:sources} (the complete sample is analyzed in Section
    \ref{sec:higher-mass}).  Note that these ``distributions of
    distributions'' are not necessarily normalized, and need not be
    shaped like the underlying model distributions.}
\end{figure}

@@

\subsection{Complete Sample}
\label{sec:higher-mass}

@@

\begin{figure}
  \begin{center}
    \plotone{plots/dist-high}
  \end{center}
  \caption{\label{fig:high-mass-dists} The median (solid line), 10\%
    (lower dashed line), and 90\% (upper dashed line) values of the
    black hole mass distribution, $p(M|\theta)$, at various masses
    implied by the posterior $p(\theta|d)$ for the models discussed in
    Sections \ref{sec:parametric-models} and
    \ref{sec:non-parametric-models}.  These distributions use the
    complete sample of 20 observations in Table \ref{tab:sources},
    including the higher-mass, wind-fed systems.  Note that these
    ``distributions of distributions'' are not necessarily normalized,
    and need not be ``shaped'' like the underlying model
    distributions.  Compare to Figure \ref{fig:dists}, which includes
    only the low-mass systems in the analysis.  Including the
    higher-mass systems tends to widen the distribution toward the
    high-mass end and, in models that allow it, produce a second,
    higher-mass peak in addition to the one in Figure
    \ref{fig:dists}. }
\end{figure}

@@

\subsubsection{Model Selection for the Complete Sample}
\label{sec:high-mass-model-selection}

@@

Reversible jump MCMC calculations of the model probabilities are
displayed in Figure \ref{fig:high-rj-evidence}; compare Figure
\ref{fig:rj}.  The relative model probabilities are given in Table
\ref{tab:rj-high}.  The exponential model is the most favored model
for the complete sample, with the two-Gaussian model the second-most
favored.  The ranking of models differs significantly from the
low-mass samples.  The improvement of the exponential model relative
to the low-mass analysis is encouraging for theoretical calculations
that attempt to model the entire population of X-ray binaries with
this mass model.  Note also that the increasing structure of the mass
distribution favors histogram models with three bins over those with
fewer bins.

@@

\begin{table}
  \begin{center}
    \begin{tabular}{|l|r|}
      \hline
      Model & Relative Evidence \\
      \hline \hline
      Exponential (Section \ref{sec:exponential}) & 0.346944 \\
      \hline
      Two Gaussian (Section \ref{sec:gaussian}) & 0.304923 \\
      \hline
      Power Law (Section \ref{sec:power-law}) & 0.120313 \\
      \hline
      Log Normal (Section \ref{sec:log-normal}) & 0.102536 \\
      \hline
      Histogram (3 Bin, Section \ref{sec:non-parametric-models}) &
      0.0473464 \\
      \hline
      Histogram (4 Bin, Section \ref{sec:non-parametric-models}) &
      0.0282086 \\
      \hline 
      Histogram (2 Bin, Section \ref{sec:non-parametric-models}) &
      0.0210994 \\
      \hline 
      Histogram (5 Bin, Section \ref{sec:non-parametric-models}) &
      0.0179703  \\
      \hline      
      Gaussian (Section \ref{sec:gaussian}) & 0.00901719 \\ 
      \hline
      Histogram (1 Bin, Section \ref{sec:non-parametric-models}) &
      0.00164214 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{\label{tab:rj-high} Relative probabilities of the various models
    from Section \ref{sec:models} implied by the complete sample of systems.  (See also Figure \ref{fig:high-rj-evidence}.)  These probabilities have been 
    computed from reversible-jump MCMC samples using the efficient jump proposal algorithm in Appendix \ref{sec:reversible-jump-mcmc}.}
\end{table}

@@

It is interesting to use our models for the underlying black hole mass
distribution in X-ray binaries to place constraints on the minimum
black hole mass.  \citet{Bailyn1998} addressed this question in the
context of a ``mass gap'' between the most massive neutron stars and
the least massive black holes.  The more recent study of
\citet{Ozel2010} also looked for a mass gap using a subset of the
models and systems presented here.  Both works found that the minimum
black hole mass is significantly above the maximum neutron star mass
\citep{Kalogera1996} of $\sim 3 \Msun$ \changed{(though
  \citet{Ozel2010} only state their evidence for a gap in terms of the
  maximum-posterior parameters for their distributions)}.

The distributions of the minimum black hole mass from the analysis of
the low-mass samples is displayed in Figure \ref{fig:min-mass}.  The
minimum black hole mass is defined as the 1\% mass quantile,
$M_{1\%}$, of the black-hole mass distribution (i.e.\ the mass lying
below 99\% of the mass distribution).  (A quantile-based definition is
necessary in the case of those distributions that do not have a hard
cutoff mass; even for those that do, like the power-law model, it can
be useful to define a ``soft'' cutoff in the event that the lower mass
hard cutoff becomes an irrelevant parameter as discussed in Section
\ref{sec:power-law}.)  \changed{For each mass distribution parameter
  sample from our MCMC, we can calculate the distribution's minimum
  black hole mass; the collection of these minimum black hole masses
  approximates the distribution of minimum black hole masses implied
  by the data in the context of that distribution.  Figure
  \ref{fig:min-mass} plots histograms of the minimum black hole mass
  samples.}

@@

\changed{We find that the best-fit model for the low-mass systems (the
  power-law) has $M_{1\%} > 4.3$ $\Msun$ in 90\% of the MCMC samples
  (i.e.\ at 90\% confidence).  This is significantly above the maximum
  theoretically-allowed neutron star mass, $\sim 3 \Msun$ (e.g.\
  \citet{Kalogera1996}).  Hence we conclude that the low-mass systems
  show strong evidence of a mass gap.

  The distribution of minimum black hole masses for the analysis of
  the complete sample (i.e.\ including the higher-mass systems) is
  shown in Figure \ref{fig:high-min-mass}.  For the most favored
  model, the exponential, we find that $M_{1\%} > 4.5$ $\Msun$ with
  90\% confidence.  We therefore conclude that there is strong
  evidence for a mass gap in the complete sample as well.}

\begin{figure}
\begin{center}
    \plotone{plots/mmin-high}
  \end{center}
  \caption{\label{fig:high-min-mass} The distributions for the minimum
    black hole mass, $M_{1\%}$, calculated from the MCMC samples for
    the models in Section \ref{sec:models} using the complete sample
    of systems.  For the two most favored models, the exponential and
    two-Gaussian, the 90\% confidence limit on the minimum black hole
    mass is 4.5 $\Msun$ and 2.3 $\Msun$, respectively.  For every
    model, we indicate the 90\% confidence bound on the minimum black
    hole mass with a vertical line.}
\end{figure}

\changed{Table \ref{tab:mmin-quants} gives the 10\%, 50\% (median),
  and 90\% quantiles for the minimum black hole mass implied by the
  low-mass sample; Table \ref{tab:mmin-quants-high} gives the same,
  but for the complete sample of systems.}

@@

\begin{table}
  \begin{center}
    \begin{tabular}{|l|l|c|r|}
      \hline
      Model & 10\% & 50\% & 90\% \\
      \hline \hline
      Exponential (Section \ref{sec:exponential}) & 4.5 & 5.4 & 6.1 \\
      \hline
      Two Gaussian (Section \ref{sec:gaussian}) & 2.3 & 4.3 & 5.5 \\
      \hline
      Power Law (Section \ref{sec:power-law}) & 5.1 & 5.9 & 6.4 \\
      \hline
      Histogram (3 Bin, Section \ref{sec:non-parametric-models}) & 4.0 & 5.5 & 6.3 \\
      \hline 
      Histogram (4 Bin, Section \ref{sec:non-parametric-models}) & 3.4 & 5.3 & 6.4 \\
      \hline
      Histogram (2 Bin, Section \ref{sec:non-parametric-models}) & 4.4 & 5.5 & 6.2 \\
      \hline
      Histogram (5 Bin, Section \ref{sec:non-parametric-models}) & 2.8 & 5.0 & 6.2 \\
      \hline
      Gaussian (Section \ref{sec:gaussian}) & -0.64 & 1.4 & 3.4 \\
      \hline 
      Histogram (1 Bin, Section \ref{sec:non-parametric-models}) & 2.9 & 4.4 & 5.5 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{\label{tab:mmin-quants-high} The 10\%, 50\% (median), and 90\% quantiles for the distribution of minimum black hole masses (in units of $\Msun$) implied by the complete sample in the context of the various models for the black hole mass distribution.  The models are listed in order of preference from model selection (Section \ref{sec:high-mass-model-selection}, Figure \ref{fig:high-rj-evidence}, and Table \ref{tab:rj-high}). }
\end{table}

>>>>

\begin{abstract}
  We perform a Bayesian analysis of the mass distribution of
  stellar-mass black holes using the observed masses of 15 low-mass
  X-ray binary systems undergoing Roche lobe overflow and five
  higher-mass, wind-fed X-ray binary systems.  Using Markov Chain
  Monte Carlo calculations, we model the mass distribution both
  parametrically---as a power law, exponential, gaussian, combination
  of two gaussians, or log-normal distribution---and
  non-parametrically---as histograms with varying numbers of bins.  We
  provide confidence bounds on the shape of the mass distribution in
  the context of each model and compare the models with each other by
  calculating their relative Bayesian evidence as supported by the
  measurements, taking into account the number of degrees of freedom
  of each model.  The mass distribution of the low-mass systems is
  best fit by a power-law, while the distribution of the combined
  sample is best fit by the exponential model.  This difference
  indicates that the low-mass subsample is not consistent with being
  drawn from the distribution of the combined population.  We examine
  the existence of a ``gap'' between the most massive neutron stars
  and the least massive black holes by considering the value,
  $M_{1\%}$, of the 1\% quantile from each black hole mass
  distribution as the lower bound of black hole masses.  Our analysis
  generates posterior distributions for $M_{1\%}$; the best model (the
  power law) fitted to the low-mass systems has a distribution of
  lower-bounds with $M_{1\%} > 4.3$ $\Msun$ with 90\% confidence,
  while the best model (the exponential) fitted to all 20 systems has
  $M_{1\%} > 4.5$ $\Msun$ with 90\% confidence.  We conclude that our
  sample of black hole masses provides strong evidence of a gap
  between the maximum neutron star mass and the lower bound on black
  hole masses.  Our results on the low-mass sample are in qualitative
  agreement with those of \citet{Ozel2010}, although our broad
  model-selection analysis more reliably reveals the best-fit
  quantitative description of the underlying mass distribution.  The
  results on the combined sample of low- and higher-mass systems are
  in qualitative agreement with \citet{Fryer2001} although the
  presence of a mass gap remains theoretically unexplained.
\end{abstract}

@@

Observations of X-ray binaries in both the optical and X-ray bands can
provide a measurement of the mass of the compact object in these
systems.  The current sample of stellar mass black holes with
dynamically measured masses includes 15 systems with low-mass, Roche
lobe overflowing donors and 5 wind-fed systems with higher-mass
donors.  Hence, sophisticated statistical analyses of the black hole
mass distribution in these systems are possible.

The first study of the mass distribution of stellar-mass black holes,
in \citet{Bailyn1998}, examined a sample of seven low-mass X-ray
binaries thought to contain a black hole, concluding in a Bayesian
analysis that the mass function was strongly peaked around seven solar
masses%
\footnote{A similar analysis of the neutron star mass distribution can
  be found in \citet{Finn1994}.}. %
\citet{Bailyn1998} found evidence of a ``gap'' between the least
massive black hole and a ``safe'' upper limit for neutron star masses
of 3 $\Msun$ (e.g.\ \citet{Kalogera1996}).  Such a gap is puzzling in
light of theoretical studies that predict a continuous distribution of
compact object supernova remnant masses with a smooth transition from
neutron stars to black holes \citep{Fryer2001}.  (We note that
\citet{Fryer2001} considered binary evolution effects only
heuristically and put forward some possible explanations for the gap
from \citet{Bailyn1998} both in the context of selection effects or in
connection to the energetics of supernova explosions.)

Towards the end of our analysis work, we became aware of a more recent
study \citep{Ozel2010}, also in a Bayesian framework, analyzing the
low-mass X-ray binary sample.  Our results are largely consistent with
those obtained by \citet{Ozel2010}, who examined 16 low-mass X-ray
binary systems containing black holes and found a strongly peaked
distribution at $7.8 \pm 1.2 \, \Msun$.  They used two models for the
mass function: a Gaussian and a decaying exponential with a minimum
``turn-on'' mass (motivated by the analytical model of the black-hole
mass function in \citet{Fryer2001}).  We note that \citet{Ozel2010} do
not provide confidence limits for the minimum black hole mass, instead
discussing only the model parameters at the peak of their posterior
distributions.  They also do not perform any model selection analysis;
thus, they give the distribution of parameters within each of their
models, but cannot say which model is more likely to correspond to the
true distribution of black hole masses.  Nevertheless, it appears that
their analysis confirms the existence of a mass gap.  \citet{Ozel2010}
discuss possible selection effects that could lead to the appearance
of a mass gap, but conclude these effects could not produce the
observed gap, which they therefore claim is a real property of the
black hole mass distribution.

@@

We define the ``minimum black hole mass'' to be the 1\% quantile,
$M_{1\%}$, in the black hole mass distribution (see Section
\ref{sec:minimum-mass}).  In qualitative agreement with
\citet{Ozel2010} and \citet{Bailyn1998}, we find strong evidence for a
mass gap among the best models for both samples.  Our analysis gives
distributions for $M_{1\%}$ implied by the data in the context of each
of our models for the black hole mass distribution.  In the context of
the best model for the low-mass systems (a power-law), the
distribution for $M_{1\%}$ gives $M_{1\%} > 4.3$ $\Msun$ with 90\%
confidence; in the context of the best model for the combined sample
of lower- and high-mass systems the distribution of $M_{1\%}$ has
$M_{1\%} > 4.5$ $\Msun$ with 90\% confidence.  Further, in the context
of models with lower evidence, most also have a mass gap, with 90\%
confidence bounds on $M_{1\%}$ significantly above a ``safe'' maximum
neutron star mass of 3 $\Msun$ \citep{Kalogera1996}.

We find that, for the low-mass X-ray binary sample, the theoretical
model from \citet{Fryer2001}---a decaying exponential---is strongly
disfavored by our model selection.  We find that the low-mass systems
are best described by a power law, followed closely by a Gaussian
(which is the second model considered by \citet{Ozel2010}).  On the
other hand, we find that the theoretical model from \citet{Fryer2001}
is the preferred model for the combined sample of low- and high-mass
X-ray binaries.  A model with two separate Gaussian peaks also has
relatively high evidence for the combined sample of systems.  The
difference in best-fitting model indicates that the low-mass subsample
is not consistent with being drawn from the distribution of the
combined population.

The structure of this paper is as follows.  In Section
\ref{sec:systems} we discuss the 15 systems that comprise the low-mass
X-ray binary black hole sample and the 5 additional higher-mass,
wind-fed systems that make up the combined sample.  In Section
\ref{sec:models} we discuss the Bayesian techniques we use to analyze
the black hole mass distribution, the techniques we use for model
selection, and the parametric and non-parametric models we will for
the black hole mass distribution.  In Section \ref{sec:results} we
discuss the results of our analysis and model selection.  In Section
\ref{sec:minimum-mass} we discuss the distribution of the minimum
black hole mass implied by the analysis of Section \ref{sec:results}.
In Section \ref{sec:conclusion} we summarize our results and comment
on the significance of the observed mass gap in the context of
theoretical models.  Appendix \ref{sec:mcmc} describes MCMC techniques
in some detail.  Appendix \ref{sec:reversible-jump-mcmc} explains our
novel algorithm for efficiently performing the reversible jump MCMC
computations used in the model comparisons of Section
\ref{sec:results} (but see also \citet{Farr2010}).

@@

The 20 X-ray binary systems on which this study is based are listed in
Table \ref{tab:sources}.  We separate the systems into 15 low-mass
systems in which the central black hole appears to be fed by
roche-lobe overflow from the secondary, and 5 higher-mass systems in
which the black hole is fed via winds (these systems all have a
secondary that appears to be more massive than the black hole).  The
low- and higher-mass systems undoubtedly have different evolutionary
tracks, and therefore it is reasonable that they would have different
black-hole mass distributions.  We will first analyze the 15 low-mass
systems alone (Section \ref{sec:results-low-mass}), and then the
combined sample of 20 systems (Section \ref{sec:higher-mass}).

@@

In this section we describe the statistical analysis we will apply to
various models for the underlying mass distribution from which the
low-mass sample and the combined sample of X-ray binary systems in
Table \ref{tab:sources} were drawn.  The results of our analysis are
presented in Section \ref{sec:results}.

@@

We consider histograms with up to five bins in this work.  We will see
that the evidence for the histogram models (see Sections
\ref{sec:model-selection}, \ref{sec:low-mass-model-selection}, and
\ref{sec:high-mass-model-selection}) from both the low-mass and
combined datasets is decreasing as the number of bins reaches five,
indicating that increasing the number of bins beyond five would not
sufficiently improve the fit to the mass distribution to compensate
for the extra parameter-space volume implied by the additional
parameters.

@@

In this section we discuss the results of our MCMC analysis of the
posterior distributions of parameters for the models in Sections
\ref{sec:parametric-models} and \ref{sec:non-parametric-models}.  We
also discuss model selection results.  The results in Section
\ref{sec:results-low-mass} apply to the low-mass sample of
systems, while those of Section \ref{sec:higher-mass} apply to the
combined sample of systems.

@@

\begin{figure}
  \begin{center}
    \plotone{plots/dist}
  \end{center}
  \caption{\label{fig:dists} The median (solid line), 10\% (lower
    dashed line) and 90\% (upper dashed line) values of the black hole
    mass distribution, $p(M|\theta)$, at various masses implied by the
    posterior $p(\theta|d)$ for the models discussed in Sections
    \ref{sec:parametric-models} and \ref{sec:non-parametric-models}.
    These distributions use only the 15 low-mass observations in Table
    \ref{tab:sources} (the combined sample is analyzed in Section
    \ref{sec:higher-mass}).  Note that these ``distributions of
    distributions'' are not necessarily normalized, and need not be
    shaped like the underlying model distributions.}
\end{figure}

@@

\subsection{Combined Sample}
\label{sec:higher-mass}

@@

\begin{figure}
  \begin{center}
    \plotone{plots/dist-high}
  \end{center}
  \caption{\label{fig:high-mass-dists} The median (solid line), 10\%
    (lower dashed line), and 90\% (upper dashed line) values of the
    black hole mass distribution, $p(M|\theta)$, at various masses
    implied by the posterior $p(\theta|d)$ for the models discussed in
    Sections \ref{sec:parametric-models} and
    \ref{sec:non-parametric-models}.  These distributions use the
    combined sample of 20 observations in Table \ref{tab:sources},
    including the higher-mass, wind-fed systems.  Note that these
    ``distributions of distributions'' are not necessarily normalized,
    and need not be ``shaped'' like the underlying model
    distributions.  Compare to Figure \ref{fig:dists}, which includes
    only the low-mass systems in the analysis.  Including the
    higher-mass systems tends to widen the distribution toward the
    high-mass end and, in models that allow it, produce a second,
    higher-mass peak in addition to the one in Figure
    \ref{fig:dists}. }
\end{figure}

@@

\subsubsection{Model Selection for the Combined Sample}
\label{sec:high-mass-model-selection}

@@

Reversible jump MCMC calculations of the model probabilities are
displayed in Figure \ref{fig:high-rj-evidence}; compare Figure
\ref{fig:rj}.  The relative model probabilities are given in Table
\ref{tab:rj-high}.  The exponential model is the most favored model
for the combined sample, with the two-Gaussian model the second-most
favored.  The ranking of models differs significantly from the
low-mass samples.  The improvement of the exponential model relative
to the low-mass analysis is encouraging for theoretical calculations
that attempt to model the entire population of X-ray binaries with
this mass model.  Note also that the increasing structure of the mass
distribution favors histogram models with three bins over those with
fewer bins.

@@

\begin{table}
  \begin{center}
    \begin{tabular}{|l|r|}
      \hline
      Model & Relative Evidence \\
      \hline \hline
      Exponential (Section \ref{sec:exponential}) & 0.346944 \\
      \hline
      Two Gaussian (Section \ref{sec:gaussian}) & 0.304923 \\
      \hline
      Power Law (Section \ref{sec:power-law}) & 0.120313 \\
      \hline
      Log Normal (Section \ref{sec:log-normal}) & 0.102536 \\
      \hline
      Histogram (3 Bin, Section \ref{sec:non-parametric-models}) &
      0.0473464 \\
      \hline
      Histogram (4 Bin, Section \ref{sec:non-parametric-models}) &
      0.0282086 \\
      \hline 
      Histogram (2 Bin, Section \ref{sec:non-parametric-models}) &
      0.0210994 \\
      \hline 
      Histogram (5 Bin, Section \ref{sec:non-parametric-models}) &
      0.0179703  \\
      \hline      
      Gaussian (Section \ref{sec:gaussian}) & 0.00901719 \\ 
      \hline
      Histogram (1 Bin, Section \ref{sec:non-parametric-models}) &
      0.00164214 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{\label{tab:rj-high} Relative probabilities of the various models
    from Section \ref{sec:models} implied by the combined sample of systems.  (See also Figure \ref{fig:high-rj-evidence}.)  These probabilities have been 
    computed from reversible-jump MCMC samples using the efficient jump proposal algorithm in Appendix \ref{sec:reversible-jump-mcmc}.}
\end{table}

@@

It is interesting to use our models for the underlying black hole mass
distribution in X-ray binaries to place constraints on the minimum
black hole mass implied by the present sample.  \citet{Bailyn1998}
addressed this question in the context of a ``mass gap'' between the
most massive neutron stars and the least massive black holes.  The
more recent study of \citet{Ozel2010} also looked for a mass gap using
a subset of the models and systems presented here.  Both works found
that the minimum black hole mass is significantly above the maximum
neutron star mass \citep{Kalogera1996} of $\sim 3 \Msun$ (though
\citet{Ozel2010} only state their evidence for a gap in terms of the
maximum-posterior parameters and not the full extent of their
distributions).

The distributions of the minimum black hole mass from the analysis of
the low-mass samples are displayed in Figure \ref{fig:min-mass}.  The
minimum black hole mass is defined as the 1\% mass quantile,
$M_{1\%}$, of the black-hole mass distribution (i.e.\ the mass lying
below 99\% of the mass distribution).  (A quantile-based definition is
necessary in the case of those distributions that do not have a hard
cutoff mass; even for those that do, like the power-law model, it can
be useful to define a ``soft'' cutoff in the event that the lower mass
hard cutoff becomes an irrelevant parameter as discussed in Section
\ref{sec:power-law}.)  For each mass distribution parameter sample
from our MCMC, we can calculate the distribution's minimum black hole
mass; the collection of these minimum black hole masses approximates
the distribution of minimum black hole masses implied by the data in
the context of that distribution.  Figure \ref{fig:min-mass} plots
histograms of the minimum black hole mass samples.

@@

We find that the best-fit model for the low-mass systems (the
power-law) has $M_{1\%} > 4.3$ $\Msun$ in 90\% of the MCMC samples
(i.e.\ at 90\% confidence).  This is significantly above the maximum
theoretically-allowed neutron star mass, $\sim 3 \Msun$ (e.g.\
\citet{Kalogera1996}).  Hence we conclude that the low-mass systems
show strong evidence of a mass gap.

The distribution of minimum black hole masses for the analysis of the
combined sample (i.e.\ including the higher-mass systems) is shown in
Figure \ref{fig:high-min-mass}.  For the most favored model, the
exponential, we find that $M_{1\%} > 4.5$ $\Msun$ with 90\%
confidence.  We therefore conclude that there is strong evidence for a
mass gap in the combined sample as well.

\begin{figure}
\begin{center}
    \plotone{plots/mmin-high}
  \end{center}
  \caption{\label{fig:high-min-mass} The distributions for the minimum
    black hole mass, $M_{1\%}$, calculated from the MCMC samples for
    the models in Section \ref{sec:models} using the combined sample
    of systems.  For the two most favored models, the exponential and
    two-Gaussian, the 90\% confidence limit on the minimum black hole
    mass is 4.5 $\Msun$ and 2.3 $\Msun$, respectively.  For every
    model, we indicate the 90\% confidence bound on the minimum black
    hole mass with a vertical line.}
\end{figure}

Table \ref{tab:mmin-quants} gives the 10\%, 50\% (median), and 90\%
quantiles for the minimum black hole mass implied by the low-mass
sample; Table \ref{tab:mmin-quants-high} gives the same, but for the
combined sample of systems.

@@

\begin{table}
  \begin{center}
    \begin{tabular}{|l|l|c|r|}
      \hline
      Model & 10\% & 50\% & 90\% \\
      \hline \hline
      Exponential (Section \ref{sec:exponential}) & 4.5 & 5.4 & 6.1 \\
      \hline
      Two Gaussian (Section \ref{sec:gaussian}) & 2.3 & 4.3 & 5.5 \\
      \hline
      Power Law (Section \ref{sec:power-law}) & 5.1 & 5.9 & 6.4 \\
      \hline
      Histogram (3 Bin, Section \ref{sec:non-parametric-models}) & 4.0 & 5.5 & 6.3 \\
      \hline 
      Histogram (4 Bin, Section \ref{sec:non-parametric-models}) & 3.4 & 5.3 & 6.4 \\
      \hline
      Histogram (2 Bin, Section \ref{sec:non-parametric-models}) & 4.4 & 5.5 & 6.2 \\
      \hline
      Histogram (5 Bin, Section \ref{sec:non-parametric-models}) & 2.8 & 5.0 & 6.2 \\
      \hline
      Gaussian (Section \ref{sec:gaussian}) & -0.64 & 1.4 & 3.4 \\
      \hline 
      Histogram (1 Bin, Section \ref{sec:non-parametric-models}) & 2.9 & 4.4 & 5.5 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{\label{tab:mmin-quants-high} The 10\%, 50\% (median), and 90\% quantiles for the distribution of minimum black hole masses (in units of $\Msun$) implied by the combined sample in the context of the various models for the black hole mass distribution.  The models are listed in order of preference from model selection (Section \ref{sec:high-mass-model-selection}, Figure \ref{fig:high-rj-evidence}, and Table \ref{tab:rj-high}). }
\end{table}
